---
title: "Download GBIF occurrences"
author:
- Damiano Oldoni
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

In this document we download (alien) species occurrences for Belgium from [GBIF](https://www.gbif.org), to feed alien species indicators and emerging species assessments.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries}
library(tidyverse)      # To do data science
library(here)           # To find files
library(rgbif)          # To use GBIF services
library(trias)          # To use functions developed for TrIAS
library(lubridate)      # To work with dates
```

# Define download query parameters

## Taxa

We don't filter on taxa in the download step as we need data at class level for compensating research effort bias (see [indicators](https://github.com/trias-project/indicators) repository) later.

## Countries

Occurrences from Belgium:

```{r define_countries}
countries <- c("BE")
```

## Basis of record

All types of occurrences, except `FOSSIL SPECIMEN` and `LIVING SPECIMEN`, which can have misleading location information (e.g. location of captive animal).

```{r define_basis_of_record}
basis_of_record <- c(
  "OBSERVATION", 
  "HUMAN_OBSERVATION",
  "MATERIAL_SAMPLE", 
  "LITERATURE", 
  "PRESERVED_SPECIMEN", 
  "UNKNOWN", 
  "MACHINE_OBSERVATION"
)
```

## Year

Occurrences with a valid year:

```{r define_year}
year_begin <- 1000
year_end <- year(Sys.Date())
```

## Geographic coordinates

Occurrences with valid geographic coordinates:

```{r define_hasCoordinate}
hasCoordinate <- TRUE
```

# Download GBIF occurrences

## Trigger download

**Note**: GBIF credentials are required in the next step. 

Trigger download:

```{r trigger_gbif_download}
# Reuse existing download (comment to trigger new download)
gbif_download_key <- "0125107-200613084148143"

# Trigger new download (commented by default)
# gbif_download_key <- occ_download(
#   pred_in("country", countries),
#   pred_in("basisOfRecord", basis_of_record),
#   pred_gte("year", year_begin),
#   pred_lte("year", year_end),
#   pred("hasCoordinate", hasCoordinate),
#   user = rstudioapi::askForPassword("GBIF username"),
#   pwd = rstudioapi::askForPassword("GBIF password"),
#   email = rstudioapi::askForPassword("Email address for notification")
# )
```

## Check status of download

```{r check_metadata}
metadata <- occ_download_meta(key = gbif_download_key)
metadata$key
metadata$status
```

Write download to list of downloads and check pending downloads:

```{r update_download_list}
update_download_list(
  file = here::here("data", "raw", "gbif_downloads.tsv"), 
  download_to_add = gbif_download_key, 
  input_checklist = ""
)
```
