---
title: "Aggregate occurrence data"
author:
- Damiano Oldoni
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

In this document we aggregate data. The goal is to produce two *data cubes*, one at class level (baseline for future corrections of research effort bias) and one at species level for taxa in the unfied checklist.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries}
library(tidyverse)      # To do datascience
library(here)           # To find files
library(rgbif)          # To use GBIF services
library(glue)           # To write queries
library(RSQLite)        # To interact with SQlite databases
```

Name and path of `.sqlite` file:

```{r name_path}
key <- "0003809-190415153152247"
sqlite_file <- paste(key, "occurrence.sqlite", sep = "_")
sqlite_path <- here::here("data", "interim", sqlite_file)
```

Table name:

```{r define_table_name}
table_name <- "occ_be"
```

Open connection to database:

```{r open_connection_to_db}
sqlite_occ <- dbConnect(SQLite(), dbname = sqlite_path)
```

# Aggregate at class level - baseline

In this section we will calculate how many occurrences have been recorded for each year, EEA cell code and class For speeding up the search in database, we create an index on these rows if not present:

```{r create_idx_class}
idx_baseline <- "idx_year_cell_class"
# get indexes on table
query <- glue_sql(
    "PRAGMA index_list({table_name})",
    table_name = table_name,
    .con = sqlite_occ
)
indexes_all <- dbGetQuery(sqlite_occ, query)

# create index if not present
if (!idx_baseline %in% indexes_all$name) {
  query <- glue_sql(
  "CREATE INDEX {`idx`} ON {table_name} ({`cols_idx`*})",
  idx = idx_baseline,
  table_name = table_name,
  cols_idx = c("year", 
               "eea_cell_code", 
               "classKey", 
               "coordinateUncertaintyInMeters"),
  .con = sqlite_occ
  )
  dbExecute(sqlite_occ, query)
}
```

Group by  `year`, `eea_cell_code` and `classKey` and count number of occurrences for each group. We also add the lowest value of coordinate uncertainty.

```{r get_datacube_class}
query <- glue_sql(
  "SELECT {`cols`*}, COUNT(_ROWID_), MIN({`coord_uncertainty`}) FROM {table} GROUP BY {`cols`*}",
  cols = c("year", "eea_cell_code", "classKey"),
  coord_uncertainty = "coordinateUncertaintyInMeters",
  table = table_name,
  .con = sqlite_occ
)
occ_cube_baseline <- 
  dbGetQuery(sqlite_occ, query) %>%
  rename(
    n = "COUNT(_ROWID_)",
    min_coord_uncertainty = "MIN(`coordinateUncertaintyInMeters`)"
)
```

Preview:

```{r preview_occ_cube_baseline}
occ_cube_baseline %>% 
  head()
```

The class names can be retrieved easily by GBIF:

```{r get_class}
class_df <- tibble(
  classKey = unique(occ_cube_baseline$classKey))
class_df <- 
  class_df %>%
  mutate(class = map_chr(
    classKey, 
    function(x) {
      name_usage(x, return = "data") %>%
        pull(scientificName)
      })
  )
class_df
```

# Aggregate at species level

## Select occurrences related to alien taxa

We extract taxa from unified checklist:

```{r get_alien_species}
datasetKey <- "6d9e952f-948c-4483-9807-575348147c7e"
alien_taxa <- name_usage(
  datasetKey = datasetKey, 
  return = "data",
  limit = 10000)  %>%
  filter(origin == "SOURCE")
```

Rank present values in checklist:

```{r overview_rank}
alien_taxa %>%
  group_by(rank) %>%
  count()
```

Taxonomic status values present in checklist:

```{r taxonStatus}
alien_taxa %>%
  group_by(taxonomicStatus) %>%
  count()
```

## Aggregate occurrences for accepted species

### Aggregate occurrences

As initial step, we start by selecting occurrences linked to taxa with rank `SPECIES` and taxonomic status `ACCEPTED` and `DOUBTFUL`:

```{r alien_taxa_species}
alien_taxa_species <- 
  alien_taxa %>%
  filter(rank == "SPECIES", 
         taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL"))
alien_taxa_species %>% head()
```

The `nubKey` of these taxa are the `speciesKey` we can filter on:

```{r get_alien_taxa_species_key}
alien_taxa_species_key <-
  alien_taxa_species %>% 
  distinct(nubKey) %>% 
  pull(nubKey)
```

Species, synonyms of that species and infraspecific taxa of that species all share the same `speciesKey`.

For speeding up the aggregation, we create an index on  `year`,  `eea_cell_code` and `speciesKey` if not present:

```{r create_idx_speciesKey}
idx_species_year_cell <- "idx_species_year_cell"
# get indexes on table
query <- glue_sql(
    "PRAGMA index_list({table_name})",
    table_name = table_name,
    .con = sqlite_occ
)
indexes_all <- dbGetQuery(sqlite_occ, query)

# create index if not present
if (!idx_species_year_cell %in% indexes_all$name) {
  query <- glue_sql(
  "CREATE INDEX {`idx`} ON {table_name} ({`cols_idx`*})",
  idx = idx_species_year_cell,
  table_name = table_name,
  cols_idx = c("year", 
               "eea_cell_code", 
               "speciesKey",
               "coordinateUncertaintyInMeters"),
  .con = sqlite_occ
  )
  dbExecute(sqlite_occ, query)
}
```

Group by `year`, `eea_cell_code` and `speciesKey`, count number of occurrences for each group and get the lowest value of coordinate uncertainty `occurrenceUncertaintyInMeters`:

```{r get_datacube_species}
query <- glue_sql(
  "SELECT {`cols`*}, COUNT(_ROWID_), MIN({`coord_uncertainty`}) FROM {table} GROUP BY {`cols`*}",
  cols = c("year", 
           "eea_cell_code", 
           "speciesKey"),
  coord_uncertainty = "coordinateUncertaintyInMeters",
  table = table_name,
  .con = sqlite_occ
)
occ_cube_species <- 
  dbGetQuery(sqlite_occ, query) %>%
  rename(
    n = "COUNT(_ROWID_)",
    min_coord_uncertainty = "MIN(`coordinateUncertaintyInMeters`)"
)
occ_cube_species <-
  occ_cube_species %>%
  mutate(speciesKey = as.integer(speciesKey)) %>%
  filter(speciesKey %in% alien_taxa_species_key)
```

Number of alien species included in occurrence cube:

```{r species_included}
paste(
  length(alien_taxa_species_key[
  which(alien_taxa_species_key %in% unique(occ_cube_species$speciesKey))]),
  "out of",
  length(alien_taxa_species_key)
)
```

Preview:

```{r preview_occ_species}
occ_cube_species %>% head()
```

### Map taxa

Grouping by `speciesKey`, we loose informations about which taxa share the same  `speciesKey`. This information could be sometimes helpful. We extract it in a separate data.frame, `taxa_species`.

First, we get all distinct taxa:

```{r get_distinct_taxa_in_occ_cube_species}
query <- glue_sql(
  "SELECT DISTINCT {`cols`*} FROM {table}",
  cols = c("speciesKey",
           "taxonKey",
           "scientificName"),
  table = table_name,
  .con = sqlite_occ
)
occ_cube_species_taxa <- 
  dbGetQuery(sqlite_occ, query) %>%
  filter(speciesKey %in% alien_taxa_species_key)
```

Some species have occurrences coming from multiple taxa:

```{r show_multiple_taxonKey}
occ_cube_species_taxa %>%
  group_by(speciesKey) %>%
  count() %>%
  filter(n > 1) %>%
  select(-n) %>%
  left_join(occ_cube_species_taxa)
```

Some species have occurrences only from taxa linked to their infraspecific taxa or synonyms. In these cases `speciesKey` is not equal to `taxonKey`:

```{r show_taxa_speciesKey_not_taxonKey}
occ_cube_species_taxa %>%
  group_by(speciesKey) %>%
  count() %>%
  rename(n_taxa = n) %>%
  left_join(occ_cube_species_taxa, by = "speciesKey") %>%
  group_by(speciesKey, n_taxa) %>%
  filter(taxonKey != speciesKey) %>%
  count() %>%
  rename(n_taxonKey_not_speciesKey = n) %>%
  filter(n_taxonKey_not_speciesKey == n_taxa) %>%
  left_join(occ_cube_species_taxa %>%
              filter(speciesKey != taxonKey),
            by = "speciesKey") %>%
  ungroup() %>%
  select(-c(n_taxa, n_taxonKey_not_speciesKey))
```

We create `taxa_species` by adding the taxonomic rank, `SPECIES`, and the taxonomic status of the species, one of `ACCEPTED` or  `DOUBTFUL`, and create a column called `include` which contains all taxa whose occurrences are linked to the species:

```{r make_taxa_species}
taxa_species <- 
  occ_cube_species_taxa %>%
  
  # get unique 'speciesKey'
  distinct(speciesKey) %>%
  
  # rename column 'speciesKey' to 'key'
  rename(key = speciesKey) %>%
  
  # GBIF query via name_usage with 'key' column as input
  pmap_dfr(name_usage, return = "data") %>%
  
  # select columns of interest
  select(speciesKey, scientificName, rank, taxonomicStatus) %>%
  
  # rename 'scientificName' to 'species_scientificName'
  rename(species_scientificName = scientificName) %>%
  
  # add these columns to original df
  right_join(occ_cube_species_taxa, by = "speciesKey") %>%
  
  # group by 'speciesKey'
  group_by(speciesKey, 
           species_scientificName,
           rank,
           taxonomicStatus) %>%
  
  # create 'includes' column
  summarize(includes = paste(
    taxonKey, 
    scientificName, 
    sep = ": ", 
    collapse = " | ")) %>%
  
  # rename 'species_scientificName' to 'scientificName'
  rename(scientificName = species_scientificName)
taxa_species
```

## Aggregate occurrences for infraspecific taxa

Accepted infraspecific taxa in unified checklist:

```{r alien_taxa_under_species}
rank_under_species <- c("SUBSPECIFICAGGREGATE",
                        "SUBSPECIES", 
                        "VARIETY",
                        "SUBVARIETY",
                        "FORM",
                        "SUBFORM"
)
alien_taxa_subspecies <-
  alien_taxa %>%
  filter(rank %in% rank_under_species, 
         taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL"))
alien_taxa_subspecies
```

For these taxa we will search by keys from field `nubKey`, the taxon keys of taxa from the GBIF taxonomy backbone.

```{r get_alien_taxa_subspecies}
alien_taxa_subspecies_key <-
  alien_taxa_subspecies %>% 
  distinct(nubKey) %>% 
  pull(nubKey)
```

Get occurrences related to these taxa and their synonyms by filtering on `acceptedTaxonKey`, group by `year`, `eea_cell_code` and `acceptedTaxonKey`, count number of occurrences and get the lowest value of coordinate uncertainty `occurrenceUncertaintyInMeters`: 

```{r get_occurrences_subspecies}
query <- glue_sql(
  "SELECT {`cols`*} FROM {table} WHERE acceptedTaxonKey IN ({subspecies_key*})",
  cols = c("year", 
           "eea_cell_code",
           "acceptedTaxonKey",
           "coordinateUncertaintyInMeters"),
  subspecies_key = alien_taxa_subspecies_key,
  table = table_name,
  .con = sqlite_occ
)
occ_cube_subspecies <- 
  dbGetQuery(sqlite_occ, query) %>%
  group_by(year, eea_cell_code, acceptedTaxonKey) %>%
  summarize(
    n = n(),
    min_coord_uncertainty = min(coordinateUncertaintyInMeters)
)
```

Number of infraspecific taxa included in `occ_cube_subspecies`:

```{r subspecies_included}
paste(
  length(alien_taxa_subspecies_key[
  which(alien_taxa_subspecies_key %in% 
          unique(occ_cube_subspecies$acceptedTaxonKey))]),
  "out of",
  length(alien_taxa_subspecies_key)
)
```

Preview:

```{r preview_occ_cube_subspecies}
occ_cube_subspecies %>% head()
```

### Map taxa

Grouping by `acceptedTaxonKey`, we loose informations about which taxa share the same  `acceptedTaxonKey`. This information could be sometimes helpful. We extract it in a separate data.frame, `taxa_subspecies`.

First, we get all distinct taxa:

```{r get_taxa_occ_cube_subspecies}
query <- glue_sql(
  "SELECT DISTINCT {`cols`*} FROM {table} 
  WHERE acceptedTaxonKey IN ({subspecies_key*})",
  cols = c("taxonKey", 
           "acceptedTaxonKey",
           "scientificName"),
  subspecies_key = alien_taxa_subspecies_key,
  table = table_name,
  .con = sqlite_occ
)
occ_cube_subspecies_taxa <- 
  dbGetQuery(sqlite_occ, query)
```

Some species have occurrences coming from multiple taxa:

```{r show_multiple_taxonKey_in_occ_cube_subspecies}
occ_cube_subspecies_taxa %>%
  group_by(acceptedTaxonKey) %>%
  count() %>%
  filter(n > 1) %>%
  select(-n) %>%
  left_join(occ_cube_subspecies_taxa)
```

Some infraspecific taxa have occurrences only from taxa linked to their synonyms. In these cases `acceptedTaxonKey` is not equal to `taxonKey`:

```{r show_taxa_acceptedTaxonKey_not_taxonKey}
occ_cube_subspecies_taxa %>%
  group_by(acceptedTaxonKey) %>%
  count() %>%
  rename(n_taxa = n) %>%
  left_join(occ_cube_subspecies_taxa, by = "acceptedTaxonKey") %>%
  group_by(acceptedTaxonKey, n_taxa) %>%
  filter(taxonKey != acceptedTaxonKey) %>%
  count() %>%
  rename(n_taxonKey_not_acceptedKey = n) %>%
  filter(n_taxonKey_not_acceptedKey == n_taxa) %>%
  left_join(occ_cube_subspecies_taxa %>%
              filter(acceptedTaxonKey != taxonKey),
            by = "acceptedTaxonKey") %>%
  ungroup() %>%
  select(-c(n_taxa, n_taxonKey_not_acceptedKey))
```

We create `taxa_subspecies` by adding:
1. taxonomic rank, one of `SUBSPECIFICAGGREGATE`, `SUBSPECIES`, `VARIETY`, `SUBVARIETY`, `FORM`, or `SUBFORM`, 
2. taxonomic status, one of `ACCEPTED` or  `DOUBTFUL`

We also create a column called `include` which contains all taxa whose occurrences are linked to the `acceptedTaxonKey`:

```{r make_taxa_species}
taxa_subspecies <- 
  occ_cube_subspecies_taxa %>%
  
  # get unique 'acceptedTaxonKey'
  distinct(acceptedTaxonKey) %>%
  
  # rename column 'acceptedTaxonKey' to 'key'
  rename(key = acceptedTaxonKey) %>%
  
  # GBIF query via name_usage with 'key' column as input
  pmap_dfr(name_usage, return = "data") %>%
  
  # rename 'scientificName' to 'accepted_scientificName'
  rename(accepted_scientificName = scientificName)

# are synonyms present?
if ("acceptedKey" %in% names(taxa_subspecies)) {
  
  taxa_subspecies <-
    taxa_subspecies %>%
    
    # populate 'acceptedKey' column for not synonyms
    mutate(acceptedKey = case_when(
      is.na(acceptedKey) ~ key,
      !is.na(acceptedKey) ~acceptedKey)
  )
} else {
  taxa_subspecies <-
    taxa_subspecies %>%
    
    # create column 'acceptedKey'
    mutate(acceptedKey = key)
}

taxa_subspecies <-
  taxa_subspecies %>%
  
  # select columns of interest
  select(acceptedKey, accepted_scientificName, rank, taxonomicStatus) %>%

  # add columns to original df
  right_join(occ_cube_subspecies_taxa, 
             by = c("acceptedKey" = "acceptedTaxonKey")) %>%
  
  # group by accepted taxon
  group_by(acceptedKey, 
           accepted_scientificName,
           rank,
           taxonomicStatus) %>%
  
  # create 'includes' column
  summarize(includes = paste(
    taxonKey, 
    scientificName, 
    sep = ": ", 
    collapse = " | ")) %>%
  
  # rename 'accepted_scientificName' to 'scientificName'
  rename(scientificName = accepted_scientificName)

taxa_subspecies
```

## Aggregate occurrences for synonyms we want to keep

Some taxa in unified checklists are synonyms. For these taxa we don't trust the link to accepted taxa provided by GBIF.

Get synonyms from unified checklist:

```{r alien_taxa_synonyms}
alien_taxa_synonyms <-
  alien_taxa %>%
  filter(!taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL"))
alien_taxa_synonyms
```

Rank distribution:

```{r rank_synonym}
alien_taxa_synonyms %>%
  group_by(rank) %>%
  count() %>%
  arrange(desc(n))
```

For these taxa we will search by keys from field `nubKey`:

```{r get_alien_taxa_synonyms_key}
alien_taxa_synonyms_key <-
  alien_taxa_synonyms %>% 
  distinct(nubKey) %>% 
  pull(nubKey)
```

Get occurrences, group by `year`, `eea_cell_code` and `taxonKey`, count number of occurrences and get the lowest value of coordinate uncertainty `occurrenceUncertaintyInMeters`: 

```{r get_occurrences_synonyms}
query <- glue_sql(
  "SELECT {`cols`*} FROM {table} WHERE taxonKey IN ({synonym_key*})",
  cols = c("year", 
           "eea_cell_code", 
           "taxonKey",
           "coordinateUncertaintyInMeters"),
  synonym_key = alien_taxa_synonyms_key,
  table = table_name,
  .con = sqlite_occ
)
occ_cube_synonym <- 
  dbGetQuery(sqlite_occ, query) %>%
  group_by(year, eea_cell_code, taxonKey) %>%
  summarize(
    n = n(),
    min_coord_uncertainty = min(coordinateUncertaintyInMeters)
)
```

Number of synonyms included in `occ_cube_synonym`:

```{r synonyms_included}
paste(
  length(alien_taxa_synonyms_key[
  which(alien_taxa_synonyms_key %in% 
          unique(occ_cube_synonym$taxonKey))]),
  "out of",
  length(alien_taxa_synonyms_key)
)
```

Preview:

```{r preview_occ_cube_subspecies}
occ_cube_synonym %>% head()
```

### Map taxa

For these taxa we don't have occcurrences linked to other related taxa. We have just to retrieve scientific name, taxonomic status and rank and make a data.frame called `taxa_synonym`. In this case the column `includes` is trivial:

```{r taxa_synonym}
taxa_synonym <- 
  
  # create a df with 'key' column
  tibble(key = alien_taxa_synonyms_key[
    which(alien_taxa_synonyms_key %in% 
            unique(occ_cube_synonym$taxonKey))]) %>%
  
  # GBIF query via name_usage with 'key' column as input
  pmap_dfr(name_usage, return = "data") %>%
  
  # select columns of interest
  select(key, scientificName, rank, taxonomicStatus) %>%
  
  # rename 'key' to 'taxonKey'
  rename(taxonKey = key) %>%
  
  # create 'includes' column
  mutate(includes = paste(
    taxonKey, 
    scientificName, 
    sep = ": ")
)
taxa_synonym
```

# Save aggregated data

## Save aggregated data at class level

Save as tab separated text file:

```{r save_baseline_datacube}
write_tsv(occ_cube_baseline,
          here::here("data", "processed", "cube_belgium_baseline.tsv"),
          na = "")
```

## Save aggregated occurrences and mapped taxa for Belgium

### Merge aggregated data

Overview of aggregated data:

```{r overview_occ_cube_species}
head(occ_cube_species)
```

```{r overview_occ_cube_subspecies}
head(occ_cube_subspecies)
```

```{r overview_occ_cube_synonym}
head(occ_cube_synonym)
```

Before merging the aggregated data.frames, we have to rename some columns.

In `occ_cube_species`, rename `speciesKey` to `taxonKey`:

```{r rename_occ_cube_species}
occ_cube_species <-
  occ_cube_species %>%
  rename(taxonKey = speciesKey)
```

In `occ_cube_subspecies`, rename `acceptedTaxonKey` to `taxonKey`:

```{r rename_occ_cube_subspecies}
occ_cube_subspecies <-
  occ_cube_subspecies %>%
  rename(taxonKey = acceptedTaxonKey)
```

Merge the three data.frames in a single *cube*, `occ_cube_belgium`:

```{r make_occ_cube_belgium}
occ_cube_belgium <- 
  occ_cube_species %>%
  bind_rows(occ_cube_subspecies) %>%
  bind_rows(occ_cube_synonym)
```

### Merge taxa

Overview of the taxa contained in the aggregated data:

```{r overview_taxa_species}
head(taxa_species)
```

```{r overview_taxa_subspecies}
head(taxa_subspecies)
```

```{r overview_taxa_synonym}
head(taxa_synonym)
```

Before merging the data.frames with taxa information, we have to rename some columns as done before with the aggregated data.frames.

In `taxa_species`, rename `speciesKey` to `taxonKey`:

```{r rename_taxa_species}
taxa_species <-
  taxa_species %>%
  rename(taxonKey = speciesKey)
```

In `taxa_subspecies`, rename `acceptedKey` to `taxonKey`:

```{r rename_taxa_subspecies}
taxa_subspecies <-
  taxa_subspecies %>%
  rename(taxonKey = acceptedKey)
```

Merge the three data.frames in a single data.frame, `taxa`:

```{r make_taxa}
taxa <- 
  taxa_species %>%
  bind_rows(taxa_subspecies) %>%
  bind_rows(taxa_synonym)
```

### Save aggregated data

Save the *occurrence cube* as tab separated text file:

```{r save_cube_belgium}
write_tsv(occ_cube_belgium, 
          here::here("data", "processed", "cube_belgium.tsv"),
          na = ""
)
```

### Save taxa

Save the taxa as tab separated text file:

```{r save_cube_belgium_taxa}
write_tsv(taxa, 
          here::here("data", "processed", "cube_belgium_taxa.tsv"),
          na = ""
)
```

Close connection:

```{r close_connection}
dbDisconnect(sqlite_occ)
```
