---
title: "Assign grid to occurrences"
author:
- Damiano Oldoni
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

In this document we assign grid to occurrences

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries}
library(tidyverse)      # To do datascience
library(sp)             # To work with geospatial data
library(here)           # To find files
library(glue)           # To write queries
library(RSQLite)        # To interact with SQlite databases
```

# Get geographic coordinates and coordinate uncertainty

Name and path of `.sqlite` file:

```{r name_path}
key <- "0044021-181108115102211"
sqlite_file <- paste(key, "occurrence.sqlite", sep = "_")
sqlite_path <- here::here("data", "interim", sqlite_file)
```

Table name:

```{r define_table_name}
table_name <- "occ_be"
```

Open connection to database:

```{r open_connection_to_db}
sqlite_occ <- dbConnect(SQLite(), dbname = sqlite_path)
```

Retrieve geographic coordinates, `decimalLatitude` and `decimalLongitude` and coordinate uncertainty, `coordinateUncertaintyInMeters`:

```{r get_geodata}
query <- glue_sql("SELECT {`cols`*} FROM {table}",
              cols = c("decimalLatitude", 
                       "decimalLongitude",
                       "coordinateUncertaintyInMeters"),
              table = table_name,
              .con = sqlite_occ
)
geodata_df <- 
  dbGetQuery(sqlite_occ, query) %>%
  as_tibble()
```

Number of occurrences:

```{r n_occs}
nrow_geodata_df <- nrow(geodata_df)
nrow_geodata_df
```

Preview:

```{r preview_geodata}
geodata_df %>% head(10)
```

Transform character columns to numbers:

```{r transform_to_numbers}
geodata_df <-
  geodata_df %>%
  mutate_if(is.character, as.double)
```

Number of occurrences per each value of `coordinateUncertaintyInMeters`:

```{r n_occ_per_uncertainty}
geodata_df %>%
  group_by(coordinateUncertaintyInMeters) %>%
  count() %>%
  arrange(desc(n))
```

We assign 1000 meters to occurrences without uncertainty:

```{r assign_fix_uncertainty}
geodata_df <- 
  geodata_df %>%
  mutate(
    coordinateUncertaintyInMeters = 
      if_else(is.na(coordinateUncertaintyInMeters),
              1000.0,
              coordinateUncertaintyInMeters)
)
```

We apply the same changes in sqlite table. To speed up the process we deactivate database synchronization and wrap the execution of the query in a transaction:

```{r assign_1000_as_uncertainty_on_db}
# speed up by deactivating db synchronization
query <- glue_sql(
  "PRAGMA synchronous = OFF",
  .con = sqlite_occ
)
dbExecute(sqlite_occ, query)

# Start transaction
dbBegin(sqlite_occ)
query <- glue_sql(
            "UPDATE {table} SET {`column`} = 1000 WHERE {`column`} = ''",
            table = table_name,
            column = "coordinateUncertaintyInMeters",
            .con = sqlite_occ)
dbExecute(sqlite_occ, query)
# Commit transaction
dbCommit(sqlite_occ)
```

# Assign grid to occurrences

We use the official grid of Belgium at 1x1km resolution as provided by EEA.

## Project geographic coordinates

We project latitude and longitude by using the projection of the grid. We transform GBIF data which have coordinate reference system equal to EPSG code 4326 to Lambert projection with EPSG code 3035:

```{r transform_to_3035}
coordinates(geodata_df) <- ~decimalLongitude+decimalLatitude
proj4string(geodata_df) <- CRS("+init=epsg:4326")
geodata_df <- spTransform(geodata_df, CRS("+init=epsg:3035"))
colnames(geodata_df@coords) <- c("x", "y")
```

## Assign occurrence within uncertainty circle

Assign the occurrence randomly within the circle with radius equal to `coordinateUncertaintyInMeters`:

```{r assign_pts_in_circle}
geodata_df@data <- 
  geodata_df@data %>% 
  mutate(random_angle = runif(nrow_geodata_df, 0, 2*pi))
geodata_df@data <- 
  geodata_df@data %>%
  mutate(random_r = runif(
    nrow_geodata_df, 0, 1) * coordinateUncertaintyInMeters)
geodata_df@data <- 
  geodata_df@data %>%
  mutate(x = geodata_df@coords[, "x"],
         y = geodata_df@coords[, "y"])
geodata_df@data <- 
  geodata_df@data %>%
  mutate(x = x + random_r * cos(random_angle),
         y = y + random_r * sin(random_angle))
geodata_df@data <- 
  geodata_df@data %>%
  select(-c(random_angle, random_r)) %>%
  select(x, y, coordinateUncertaintyInMeters)
```

Preview:

```{r preview_geodata_df_data}
geodata_df@data %>% head(n = 10)
```

Where `x` and `y` are the new coordinates while in `@coords` we have the original coordinates:

```{r preview_geodata_df_coords}
geodata_df@coords[1:10,]
```

## Assign occurrences to grid cells

We assign each occurrence to a grid cell. 

```{r asign_grid}
geodata_df@data <- 
  geodata_df@data %>%
  mutate(eea_cell_code = paste0("1km", 
                                  "E", floor(x/1000), 
                                  "N", floor(y/1000)))
```
Preview:

```{r preview_gridcell}
geodata_df@data %>% head(n = 10)
```

## Add grid cell code to sqlite file

We can now add the column `eea_cell_code` to the table `occ_be` of sqlite file.
We first create the new column  `eaa_cell_code` in the table:

```{r add_eaa_cellcode_to_sqlite}
new_col <- "eea_cell_code"
query <- glue_sql("ALTER TABLE {table} ADD COLUMN {colname} {type}",
                  table = table_name,
                  colname = new_col,
                  type = "CHARACTER",
                  .con = sqlite_occ
)
dbExecute(sqlite_occ, query)
```

And then we populate it with the values in `geodata_df@data$eaa_cell_code`.

```{r add_values_cellcode}
# Start transaction
dbBegin(sqlite_occ)
dbExecute(sqlite_occ,
          glue_sql(
            "UPDATE {table} SET {`column`} = :eea_cell_code WHERE _ROWID_ = :id",
            table = table_name,
            column = new_col,
            .con = sqlite_occ),
          params = data.frame(eea_cell_code = geodata_df@data$eea_cell_code,
                              id = rownames(geodata_df@data)))
# Commit transaction
dbCommit(sqlite_occ)
```

Preview:

```{r}
query <- glue_sql("SELECT * FROM {table} WHERE _ROWID_ <= 10",
                  table = table_name,
                  .con = sqlite_occ)
dbGetQuery(sqlite_occ, query)
```

Close connection:

```{r close_connection}
dbDisconnect(sqlite_occ)
```
