---
title: "Download GBIF occurrences"
author:
- Damiano Oldoni
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to retrieve occurrences for assessing the emerging status of alien species in Belgium.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(tidyverse)
library(lubridate)
# to find files
library(here)
# GBIF package
library(rgbif)
# Project package
library(trias)
```

Define the file where all downloads we've triggered are added:

```{r set_file_paths}
downloads_output_file <- here("data", "processed", "gbif_downloads.tsv")
```

# Define download query parameters

## Define taxa

As the list of alien species from the unified checklist is too long for a query to the GBIF API, we don't define any taxa:

```{r define_taxa}
taxa_input_file <- ""
taxon_keys <- c()
```

## Define countries

We will limit our occurrences search for specific countries. To allow better tracking, this filter will also be added to `download_output_file`:

```{r define_countries}
countries <- c("BE")
```

## Define basis of record

Everything except `FOSSIL SPECIMEN` and `LIVING SPECIMEN`:

```{r define_basis_of_record}
basis_of_record <- c(
  "OBSERVATION", 
  "HUMAN_OBSERVATION",
  "MATERIAL_SAMPLE", 
  "LITERATURE", 
  "PRESERVED_SPECIMEN", 
  "UNKNOWN", 
  "MACHINE_OBSERVATION"
)
```

## Define year

We will limit to occurrence with valid year.

```{r define_year}
year_begin <- 1000
year_end <- year(Sys.Date())
```

## Define geographic coordinates

We need occurrences with valid geographic coordinates:

```{r define_hasCoordinate}
hasCoordinate <- TRUE
```

## Download GBIF occurrences

**Note**: GBIF credentials are required in the next step.

Trigger download:

```{r trigger_gbif_download}
# Important: query arguments need to be pasts as strings of format "key = value"
gbif_download_key <- occ_download(
  paste0("country = ", paste(countries, collapse = ",")),
  paste0("basisOfRecord = ", paste(basis_of_record, collapse = ",")),
  paste0("year >= ", year_begin),
  paste0("year <= ", year_end),
  paste0("hasCoordinate = ", hasCoordinate),
  curlopts = list(verbose = TRUE),
  # user = rstudioapi::askForPassword("GBIF username"),
  # pwd = rstudioapi::askForPassword("GBIF password"),
  # email = rstudioapi::askForPassword("Email address for notification")
  user = Sys.getenv("gbif_user"),
  pwd = Sys.getenv("gbif_pwd"),
  email = Sys.getenv("gbif_email")
)
```

If you want to run it locally, set GBIF username, password and e-mail in your `.Renviron` file as `gbif_user`, `gbif_pwd`, `gbif_email` respectively and run the following code:

```r
gbif_download_key <- occ_download(
  paste0("country = ", paste(countries, collapse = ",")),
  paste0("basisOfRecord = ", paste(basis_of_record, collapse = ",")),
  paste0("year >= ", year_begin),
  paste0("year <= ", year_end),
  paste0("hasCoordinate = ", hasCoordinate),
  curlopts = list(verbose = TRUE),
  # user = rstudioapi::askForPassword("GBIF username"),
  # pwd = rstudioapi::askForPassword("GBIF password"),
  # email = rstudioapi::askForPassword("Email address for notification")
  user = Sys.getenv("gbif_user"),
  pwd = Sys.getenv("gbif_pwd"),
  email = Sys.getenv("gbif_email")
)
```

For more details, type `?occ_download` in your R console.

## Check status of download:

```{r}
metadata <- occ_download_meta(key = gbif_download_key)
metadata$key
metadata$status
```

Write download to list of downloads and check pending downloads:

```{r}
update_download_list(
  file = downloads_output_file, 
  download_to_add = gbif_download_key, 
  input_checklist = taxa_input_file
)
```
