---
title: "Aggregate occurrence data"
author:
- Damiano Oldoni
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

In this document we aggregate data. The goal is to produce a *data cube* at species level. The dimensions are:

1. taxon
2. year
3. geographic cell code

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries_eu}
library(tidyverse)      # To do datascience
library(sp)             # To work with geospatial data
library(here)           # To find files
library(rgbif)          # To use GBIF services
library(glue)           # To write queries
```

# Read data

## Read modelling species data

Read taxa for modelling from  `modelling_species.tsv`:

```{r read_modelling_taxa}
modelling_taxa <- read_tsv(
  here::here("references", "modelling_species.tsv"),
  na = "",
  quote = "",
  col_types = cols(
    .default = col_character(),
    backbone_taxonKey = col_integer()
  ))
```

Number of taxa:

```{r nrow_modelling_taxa}
nrow(modelling_taxa)
```

Preview:

```{r preview_modelling_taxa}
modelling_taxa %>% head()
```

## Read occurrence data

Name and path of occurrence data file:

```{r name_path_eu}
key <- "0004487-200127171203522"
occ_file <- paste(key, "occurrence.tsv", sep = "_")
occ_path <- here::here("data", "interim", occ_file)
```

### Define column type specifications

Name of columns:

```{r get_cols_occs_eu}
cols_occ_file <- read_delim(
  occ_path, "\t", n_max = 1,
  quote = ""
)
cols_occ_file <- names(cols_occ_file)
```

The following columns should contain integers:

1. `*Key`, e.g. `taxonKey`, `speciesKey`
2. `*DayOfYear`: `startDayOfYear` and  `endDayOfYear`  
3. `year`
4. `month`
5. `day`

```{r define_col_integer_for_specific_columns_eu}
int_colnames <- 
  cols_occ_file[str_detect(cols_occ_file, "Key") & 
                !str_detect(cols_occ_file, "datasetKey")]
int_colnames <- c(
  int_colnames,
  cols_occ_file[str_detect(cols_occ_file, "DayOfYear")],
  cols_occ_file[cols_occ_file == "year"],
  cols_occ_file[cols_occ_file == "month"],
  cols_occ_file[cols_occ_file == "day"]
)

int_cols <- 
  map(int_colnames, ~ col_integer()) %>% 
  setNames(int_colnames)
```

The following columns should contain real numbers:

1. `decimal*`: `decimalLatitude` and `decimalLongitude`
2. `coordinate*`: `coordinateUncertaintyInMeters` and `coordinatePrecision`
3. `pointRadiusSpatialFit`

```{r define_col_double_for_specific_columns_eu}
real_colnames <- cols_occ_file[str_detect(cols_occ_file, "decimal")]
real_colnames <- c(
  real_colnames,
  cols_occ_file[str_detect(cols_occ_file, "coordinate")],
  cols_occ_file[cols_occ_file == "pointRadiusSpatialFit"]
)

real_cols <- 
  map(real_colnames, ~ col_double()) %>% 
  setNames(real_colnames)
```

The other columns contain text:

```{r define_col_text_for_other_columns_eu}
char_colnames <- cols_occ_file[!cols_occ_file %in% real_colnames & 
                               !cols_occ_file %in% int_colnames]
char_cols <- 
  map(char_colnames, ~ col_character()) %>% 
  setNames(char_colnames)
```

Final column specification:

```{r cols_type_to_use_eu}
col_specs <- cols()
col_specs$cols <- c(char_cols, int_cols, real_cols)
col_specs
```

### Import occurrence data with assigned cell code

Import occurrence data:

```{r read_occs_in_eu}
occ_eu <- read_tsv(
  here::here("data", 
             "interim", 
             paste0(key,"_occurrence.tsv")),
  na = "",
  quote = "",
  col_types = col_specs)
```

# Aggregate at species level

## Get taxonomic status and rank of modelling species

We add taxonomic status and rank to `modelling_taxa`. This information comes from GBIF Taxonomy Backbone:

```{r get_status_rank_modelling_taxa}
modelling_taxa <- 
  modelling_taxa %>%
  bind_cols(
    map_dfr(modelling_taxa$backbone_taxonKey,
        function(taxon) {
          name_usage(taxon, 
                     return = "data")}) %>%
      select(key, taxonomicStatus, rank, speciesKey)
)
```

Rank values of modelling taxa:

```{r overview_rank_modelling_taxa_eu}
modelling_taxa %>%
  group_by(rank) %>%
  count()
```

Taxonomic status of modelling taxa:

```{r overview_taxonStatus__modelling_taxa_eu}
modelling_taxa %>%
  group_by(taxonomicStatus) %>%
  count()
```

## Aggregate occurrences for accepted species

Taxa with rank `SPECIES` and taxonomic status `ACCEPTED` or `DOUBTFUL`:

```{r modelling_taxa_species_accepted_eu}
modelling_taxa_species <- 
  modelling_taxa %>%
  filter(rank == "SPECIES", 
         taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL"))
modelling_taxa_species
```

The `speciesKey` of these taxa will be used to filter occurrence data:

```{r get_modelling_taxa_species_key}
modelling_taxa_species_key <-
  modelling_taxa_species %>% 
  distinct(speciesKey) %>% 
  pull(speciesKey)
```

### Aggregate occurrences

Species, synonyms of those species and infraspecific taxa of those species all share the same `speciesKey`.

Group by `year`, `eea_cell_code` and `speciesKey`, count number of occurrences for each group and get the lowest value of coordinate uncertainty `occurrenceUncertaintyInMeters`:

```{r get_datacube_species_eu}
occ_cube_eu_species <- 
  occ_eu %>%
  group_by(year, eea_cell_code, speciesKey) %>%
  summarize(n = n(),
            min_coord_uncertainty = min(coordinateUncertaintyInMeters)) %>%
  ungroup() %>%
  filter(speciesKey %in% modelling_taxa_species_key)
```

Number of modelling species included in `occ_cube_eu_species`:

```{r species_included_eu}
paste(length(modelling_taxa_species_key[
  which(modelling_taxa_species_key %in% unique(occ_cube_eu_species$speciesKey))
]), "out of", length(modelling_taxa_species_key))
```

Preview:

```{r preview_occ_species_eu}
occ_cube_eu_species %>% head()
```

### Map taxa

Grouping by `speciesKey`, we loose informations about which taxa share the same  `speciesKey`. This information could be sometimes helpful. We extract it in a separate data.frame, `taxa_species_eu`.

First, we get all distinct taxa:

```{r get_distinct_taxa_in_occ_cube_species_eu}
occ_cube_eu_species_taxa <- 
  occ_eu %>%
  distinct(speciesKey, taxonKey, scientificName) %>%
  filter(speciesKey %in% modelling_taxa_species_key)
```

Some species have occurrences coming from multiple taxa:

```{r show_multiple_taxonKey_eu}
occ_cube_eu_species_taxa %>%
  group_by(speciesKey) %>%
  count() %>%
  filter(n > 1) %>%
  select(-n) %>%
  left_join(occ_cube_eu_species_taxa)
```

Some species have occurrences only from taxa linked to their infraspecific taxa or synonyms. In these cases `speciesKey` is not equal to `taxonKey`:

```{r show_taxa_speciesKey_not_taxonKey_eu}
occ_cube_eu_species_taxa %>%
  group_by(speciesKey) %>%
  count() %>%
  rename(n_taxa = n) %>%
  left_join(occ_cube_eu_species_taxa, by = "speciesKey") %>%
  group_by(speciesKey, n_taxa) %>%
  filter(taxonKey != speciesKey) %>%
  count() %>%
  rename(n_taxonKey_not_speciesKey = n) %>%
  filter(n_taxonKey_not_speciesKey == n_taxa) %>%
  left_join(occ_cube_eu_species_taxa %>%
              filter(speciesKey != taxonKey),
            by = "speciesKey") %>%
  ungroup() %>%
  select(-c(n_taxa, n_taxonKey_not_speciesKey))
```

We create `taxa_species_eu` by adding the taxonomic rank, `SPECIES`, and the taxonomic status of the species, one of `ACCEPTED` or  `DOUBTFUL`, and create a column called `include` which contains all taxa whose occurrences are linked to the species:

```{r make_taxa_species_eu}
taxa_species_eu <- 
  occ_cube_eu_species_taxa %>%
  
  # get unique 'speciesKey'
  distinct(speciesKey) %>%
  
  # rename column 'speciesKey' to 'key'
  rename(key = speciesKey) %>%
  
  # GBIF query via name_usage with 'key' column as input
  pmap_dfr(name_usage, return = "data") %>%
  
  # select columns of interest
  select(speciesKey, scientificName, rank, taxonomicStatus) %>%
  
  # rename 'scientificName' to 'species_scientificName'
  rename(species_scientificName = scientificName) %>%
  
  # add these columns to original df
  right_join(occ_cube_eu_species_taxa, by = "speciesKey") %>%
  
  # group by 'speciesKey'
  group_by(speciesKey, 
           species_scientificName,
           rank,
           taxonomicStatus) %>%
  
  # create 'includes' column
  summarize(includes = paste(
    taxonKey, 
    scientificName, 
    sep = ": ", 
    collapse = " | ")) %>%
  
  # rename 'species_scientificName' to 'scientificName'
  rename(scientificName = species_scientificName)
taxa_species_eu
```

## Aggregate occurrences for infraspecific taxa

Accepted infraspecific taxa in unified checklist:

```{r modelling_taxa_under_species_eu}
rank_under_species <- c("SUBSPECIFICAGGREGATE",
                        "SUBSPECIES", 
                        "VARIETY",
                        "SUBVARIETY",
                        "FORM",
                        "SUBFORM"
)
modelling_taxa_subspecies <-
  modelling_taxa %>%
  filter(rank %in% rank_under_species, 
         taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL"))
modelling_taxa_subspecies
```

For these taxa we will search by keys from field `key`, as the taxon keys come from the GBIF taxonomy backbone.

```{r get_modelling_taxa_subspecies_key_eu}
modelling_taxa_subspecies_key <-
  modelling_taxa_subspecies %>% 
  distinct(key) %>% 
  pull(key)
```

Get occurrences related to these taxa and their synonyms by filtering on `acceptedTaxonKey`, group by `year`, `eea_cell_code` and `acceptedTaxonKey`, count number of occurrences and get the lowest value of coordinate uncertainty `occurrenceUncertaintyInMeters`: 

```{r get_occurrences_subspecies_eu}
occ_cube_eu_subspecies <- 
  occ_eu %>%
  filter(acceptedTaxonKey %in% modelling_taxa_subspecies_key) %>%
  select(year, 
         eea_cell_code, 
         speciesKey, 
         acceptedTaxonKey, 
         coordinateUncertaintyInMeters) %>%
  group_by(year, eea_cell_code, speciesKey, acceptedTaxonKey) %>%
  summarize(
    n = n(),
    min_coord_uncertainty = min(coordinateUncertaintyInMeters)
)
```

Number of infraspecific taxa included in `occ_cube_subspecies`:

```{r subspecies_included_eu}
paste(length(modelling_taxa_subspecies_key[
  which(modelling_taxa_subspecies_key %in% 
          unique(occ_cube_eu_subspecies$acceptedTaxonKey))]),
  "out of",
  length(modelling_taxa_subspecies_key))
```

Preview:

```{r preview_occ_cube_eu_subspecies}
occ_cube_eu_subspecies %>% head()
```

### Map taxa

Grouping by `acceptedTaxonKey`, we loose informations about which taxa share the same  `acceptedTaxonKey`. This information could be sometimes helpful. We extract it in a separate data.frame, `taxa_subspecies_eu`.

First, we get all distinct taxa:

```{r get_taxa_occ_cube_eu_subspecies}
occ_cube_eu_subspecies_taxa <- 
  occ_eu %>%
  filter(acceptedTaxonKey %in% modelling_taxa_subspecies_key) %>%
  distinct(taxonKey,
           acceptedTaxonKey,
           scientificName
)
```

Some species have occurrences coming from multiple taxa:

```{r show_multiple_taxonKey_eu}
occ_cube_eu_subspecies_taxa %>%
  group_by(acceptedTaxonKey) %>%
  count() %>%
  filter(n > 1) %>%
  select(-n) %>%
  left_join(occ_cube_eu_subspecies_taxa)
```

Some infraspecific taxa could have occurrences only from taxa linked to their synonyms. In these cases `acceptedTaxonKey` is not equal to `taxonKey`:

```{r show_taxa_speciesKey_not_taxonKey_eu}
occ_cube_eu_subspecies_taxa %>%
  group_by(acceptedTaxonKey) %>%
  count() %>%
  rename(n_taxa = n) %>%
  left_join(occ_cube_eu_subspecies_taxa, by = "acceptedTaxonKey") %>%
  group_by(acceptedTaxonKey, n_taxa) %>%
  filter(taxonKey != acceptedTaxonKey) %>%
  count() %>%
  rename(n_taxonKey_not_acceptedKey = n) %>%
  filter(n_taxonKey_not_acceptedKey == n_taxa) %>%
  left_join(occ_cube_eu_subspecies_taxa %>%
              filter(acceptedTaxonKey != taxonKey),
            by = "acceptedTaxonKey") %>%
  ungroup() %>%
  select(-c(n_taxa, n_taxonKey_not_acceptedKey))
```

We create `taxa_subspecies_eu` by adding the taxonomic rank, one of `SUBSPECIFICAGGREGATE`, `SUBSPECIES`, `VARIETY`, `SUBVARIETY`, `FORM`, or `SUBFORM`, and the taxonomic status, one of `ACCEPTED` or  `DOUBTFUL`, and create a column called `include` which contains all taxa whose occurrences are linked to the `acceptedTaxonKey`:

```{r make_taxa_species_eu}
taxa_subspecies_eu <- 
  occ_cube_eu_subspecies_taxa %>%
  
  # get unique 'acceptedTaxonKey'
  distinct(acceptedTaxonKey) %>%
  
  # rename column 'acceptedTaxonKey' to 'key'
  rename(key = acceptedTaxonKey) %>%
  
  # GBIF query via name_usage with 'key' column as input
  pmap_dfr(name_usage, return = "data")

if (ncol(taxa_subspecies_eu) > 0) {
  taxa_subspecies_eu <-
    taxa_subspecies_eu %>%

        # rename 'scientificName' to 'accepted_scientificName'
    rename(accepted_scientificName = scientificName)

  # are synonyms present?
  if ("acceptedKey" %in% names(taxa_subspecies_eu)) {
    
    taxa_subspecies_eu <-
      taxa_subspecies_eu %>%
      
      # populate 'acceptedKey' column for not synonyms
      mutate(acceptedKey = case_when(
        is.na(acceptedKey) ~ key,
        !is.na(acceptedKey) ~acceptedKey)
    )
  } else {
    taxa_subspecies_eu <-
      taxa_subspecies_eu %>%
      
      # create column 'acceptedKey'
      mutate(acceptedKey = key)
  }
  taxa_subspecies_eu <-
  taxa_subspecies_eu %>%
  
  # select columns of interest
  select(acceptedKey, accepted_scientificName, rank, taxonomicStatus) %>%

  # add columns to original df
  right_join(occ_cube_eu_subspecies_taxa, 
             by = c("acceptedKey" = "acceptedTaxonKey")) %>%
  
  # group by accepted taxon
  group_by(acceptedKey, 
           accepted_scientificName,
           rank,
           taxonomicStatus) %>%
  
  # create 'includes' column
  summarize(includes = paste(
    taxonKey, 
    scientificName, 
    sep = ": ", 
    collapse = " | ")) %>%
  
  # rename 'accepted_scientificName' to 'scientificName'
  rename(scientificName = accepted_scientificName)
}

taxa_subspecies_eu
```

## Aggregate occurrences for synonyms we want to keep

Some taxa in unified checklists are synonyms. For these taxa we don't trust the link to accepted taxa provided by GBIF.

Get synonyms from unified checklist:

```{r modelling_taxa_synonyms_eu}
modelling_taxa_synonyms <-
  modelling_taxa %>%
  filter(!taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL"))
modelling_taxa_synonyms
```

Rank distribution:

```{r rank_synonym_eu}
modelling_taxa_synonyms %>%
  group_by(rank) %>%
  count() %>%
  arrange(desc(n))
```

For these taxa we will search by keys from field `key`, the key from GBIF Taxonomy Backbone:

```{r get_modelling_taxa_synonyms_key_eu}
modelling_taxa_synonyms_key <-
  modelling_taxa_synonyms %>% 
  distinct(key) %>% 
  pull(key)
```

Get occurrences, group by `year`, `eea_cell_code` and `taxonKey`, count number of occurrences and get the lowest value of coordinate uncertainty `occurrenceUncertaintyInMeters`: 

```{r get_occurrences_synonyms_eu}
occ_cube_eu_synonym <- 
  occ_eu %>%
  filter(taxonKey %in% modelling_taxa_synonyms_key) %>%
  select(year, 
         eea_cell_code, 
         taxonKey, 
         coordinateUncertaintyInMeters) %>%
  group_by(year, eea_cell_code, taxonKey) %>%
  summarize(
    n = n(),
    min_coord_uncertainty = min(coordinateUncertaintyInMeters)
)
```

Number of synonyms included in `occ_cube_synonym`:

```{r synonyms_included_eu}
paste(length(modelling_taxa_synonyms_key[
  which(modelling_taxa_synonyms_key %in% 
          unique(occ_cube_eu_synonym$taxonKey))]),
  "out of",
  length(modelling_taxa_synonyms_key))
```


Preview:

```{r preview_occ_cube_eu_synonym}
occ_cube_eu_synonym %>% head()
```

### Map taxa

For these taxa we don't have occcurrences linked to other related taxa. We have just to retrieve scientific name, taxonomic status and rank. We extract this information and make a data.frame called `taxa_synonym_eu`. In this case the column `includes` is trivial:

```{r taxa_synonym_eu}
taxa_synonym_eu <- 
  
  # create a df with 'key' column
  tibble(key = modelling_taxa_synonyms_key[
    which(modelling_taxa_synonyms_key %in% 
            unique(occ_cube_eu_synonym$taxonKey))]) %>%
  
  # GBIF query via name_usage with 'key' column as input
  pmap_dfr(name_usage, return = "data") %>%
  
  # select columns of interest
  select(key, scientificName, rank, taxonomicStatus) %>%
  
  # rename 'key' to 'taxonKey'
  rename(taxonKey = key) %>%
  
  # create 'includes' column
  mutate(includes = paste(
    taxonKey, 
    scientificName, 
    sep = ": ")
)
taxa_synonym_eu
```

# Save aggregated data

## Save aggregated occurrences and mapped taxa for Europe

### Merge aggregated data

Overview of aggregated data:

```{r overview_occ_cube_eu_species}
head(occ_cube_eu_species)
```

```{r overview_occ_cube_eu_subspecies}
head(occ_cube_eu_subspecies)
```

```{r overview_occ_cube_eu_synonym}
head(occ_cube_eu_synonym)
```

Before merging the aggregated data.frames, we have to rename some columns.

In `occ_cube_eu_species`, rename `speciesKey` to `taxonKey`:

```{r rename_occ_cube_eu_species}
occ_cube_eu_species <-
  occ_cube_eu_species %>%
  rename(taxonKey = speciesKey)
```

In `occ_cube_eu_subspecies`, if not empty, rename `acceptedTaxonKey` to `taxonKey`:

```{r rename_occ_cube_eu_subspecies}
if (ncol(occ_cube_eu_subspecies) > 0) {
  occ_cube_eu_subspecies <-
    occ_cube_eu_subspecies %>%
    rename(taxonKey = acceptedTaxonKey)
}
```

Merge the three data.frames in a single *cube*, ` eu_modellingtaxa_cube`:

```{r make_ eu_modellingtaxa_cube}
eu_modellingtaxa_cube <- 
  occ_cube_eu_species %>%
  bind_rows(occ_cube_eu_subspecies) %>%
  bind_rows(occ_cube_eu_synonym)
```

### Merge taxa

Overview of the taxa contained in the aggregated data:

```{r overview_taxa_species_eu}
head(taxa_species_eu)
```

```{r overview_taxa_subspecies_eu}
head(taxa_subspecies_eu)
```

```{r overview_taxa_synonym_eu}
head(taxa_synonym_eu)
```

Before merging the data.frames, we have to rename some columns as done before with the aggregated data.frames.

In `taxa_species_eu`, rename `speciesKey` to `taxonKey`:

```{r rename_taxa_species_eu}
taxa_species_eu <-
  taxa_species_eu %>%
  rename(taxonKey = speciesKey)
```

In `taxa_subspecies_eu`, if not empty, rename `acceptedKey` to `taxonKey`:

```{r rename_taxa_subspecies_eu}
if (ncol(taxa_subspecies_eu) > 0) {
  taxa_subspecies_eu <-
    taxa_subspecies_eu %>%
    rename(taxonKey = acceptedKey)
}
```

Merge the three data.frames in a single data.frame, ` eu_modellingtaxa_info`:

```{r make_eu_modellingtaxa_cube}
eu_modellingtaxa_info <- 
  taxa_species_eu %>%
  bind_rows(taxa_subspecies_eu) %>%
  bind_rows(taxa_synonym_eu)
```

### Save aggregated data

Save the *occurrence cube* as comma separated text file:

```{r save_eu_modellingtaxa_cube}
write_csv(eu_modellingtaxa_cube, 
          here::here("data", "processed", "eu_modellingtaxa_cube.csv"),
          na = ""
)
```

### Save taxa

Save the taxa as comma separated text file:

```{r save_eu_modellingtaxa_info}
write_csv(eu_modellingtaxa_info, 
          here::here("data", "processed", "eu_modellingtaxa_info.csv"),
          na = ""
)
```
